Minimizing Time Required For Graduation

### 2010-12-15 paper/paper.tex
Final revisions, more spelling and typograpical fixes, new assumptions section

\subsection{Representing Prerequisites} Initially, we represented course
prerequisites using a directed acyclic graph.  Each node represented a course, and edges
existed between nodes $i$ and $j$ if and only if $i$ was a prerequisite for $j$.
This model works well if each course has only one prerequisite.  However,
this model fails to represent courses that have multiple prerequisites
or courses that have alternative prerequisites.  This is the case for some courses.
We therefore concluded that it is infeasible to model prerequisites using a graph.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{prereq_logic_example} 
    \end{center}
    \caption{Evaluation of a rule} 
    \label{logic_ex} 
\end{figure}

@@

Instead of trying to solve the complete problem we chose to use the "rolling
horizon approach" used by Xiangtong et al \cite{xiangton:informs} when
scheduling classes for airline pilots.  Xiangtong et al. also could not develop
an optimal class schedule in one pass, instead they picked optimal classes
using a heuristic over a well demarcated time period.  Once they had a solution
to the smaller time period, they increased the size of the time period, or rolled
the horizon, and used the previously optimal solution to help determine
a solution to the new larger problem.  The team's algorithm continued rolling the
time horizon on the problem until they had a solution to the entire problem.

@@

@@

\subsubsection{Procedure} We present the procedure for populating the set $L$ as
follows: \begin{enumerate} \item For each rule $r \in R$ we iterate over each
literal $p$ in the rule.  Recall that $p$ represents a course, for example
$MATH125$.  We then check if $p \in T$.  If $p \in T$, we say $p$ evaluates
to true; otherwise, $p$ evaluates to false.  Using rules of logic we evaluate
$r$.  If $r$ is true, we add the course associated with it to $L$.  Once this
is complete, $L$ will consist of every course a student is eligible to take, based solely
on satisfied prerequisites.  \item We now prune $L$ by removing all
courses that are not offered in the current quarter.  $L$ now contains
only classes that we can take given the current quarter and the classes we have
already taken.  \item We now construct a conflict graph, $G_{conflict}$, where
every course offering in $L$ is a node, and there is an edge between two nodes
$i$ and $j$ if and only if the time of course offering $i$ overlaps the time of
course offering $j$.  Figure~\ref{conflictg} shows an example conflict graph for
one time period.  \end{enumerate}
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.35]{conflict_graph_ex} 
    \end{center} 
    \caption{A Conflict Graph} 
    \label{conflictg} 
\end{figure}

\subsubsection{Analysis} The selection procedure involves a lot of set queries
and iteration, but in asymptotic running time the algorithm is efficient.
Let $n$ be the total number of classes ever available to take.
In the worst case, $P$ has a complex rule for all $n$
courses, and checking each rule in $P$ is linear in the number of propositions.
However, we know {\it a priori} from analysis of our data set that the largest
rule has twelve propositions, so we can bound the rule check time by a constant,
$C$, and therefore step one runs in $O(n)$ time.

@@

Additionally, we know that not every course is offered every quarter.  If
a course is not offered until the following year, we could potentially add an
entire year to the schedule.  This is a situation we hope to avoid.  Our heuristic
function must therefore consider how often and how far in the future a given
course is offered.  For example, if we are deciding between two course offering
that are identical except one is offered once a year, and the other is offered
three times a year, we want to take the course only offered once a year, as we
won't have the opportunity to take it for another entire year.
\begin{figure} [ht]
    \begin{center}
        \includegraphics[scale=0.35]{more_prereq_tree}
    \end{center}
    \caption{A Prerequisite DAG}
    \label{prereq}
\end{figure}

We these considerations in mind, we begin to develop our heuristic function.  In
order to define our heuristic function, we must first define some auxiliary
functions on course offering.  \begin{itemize} \item Let $TimeFromNow(c, q)$ be
difference in quarters between the current quarter, $q$, and the next quarter
a given course offering, $c$, is offered.  As there are four quarters, the range
of this function is $\{0, 1, 2, 3\}$.  \item Let $Quarters(c)$ be the number of
quarters per year a given course, $c$, is offered.  As there are four quarters,
the range of this function is $\{0, 1, 2, 3, 4\}$.  \item Let $Dependents(c)$ be
the number of course that either directly or indirectly require a given course,
$c$.  In Figure~\ref{prereq}, $Dependents(MATH125)$ = 6 while
$Dependents(MATH307)$ = 1 \end{itemize}

@@

Where $c$ is a given course offering, $q$ is the current quarter, and $\alpha$,
$\beta$, and $\gamma$ are real numbers, giving a weight to piece of the
heuristic.  We determine these scalars experimentally.

\subsubsection{Determining $Dependents(c)$} We briefly address how to determine
the dependent value of a particular course. Figure~\ref{prereq} indicates the
method for determining the dependent value of a course; generate the graph of
prerequisites and then count the number of children in the subtree rooted at
a particular course node. If this was the only consideration though, we could
potentially assign a very high value to a course that is not related to
a student's major becuase that course had a large number of dependents. The
purpose of our system is to provide schedules that get a student through a major
as quickly as possible, so we should only value courses that lead to graduation.
The set $S$ describes the classes a student will need to graduate.

To assign values to courses, we start with the courses in $S$ and work backwards
using the prerequisite rules in $R$ to determine how to traverse the
prerequisite DAG. Traversing the graph using this method, we determine the value
of the course by counting the children in trees rooted at the current node in
the traversal. By starting with the courses in $S$ we ensure that we only given
values to courses that we need to take, instead of any courses with a large
number of dependents.

@@

\begin{equation}
   X_i + X_j \leq 1, \forall \text{ pairs } (X_i,X_j) \text{ where } X_i \text{and} X_j \text{ are offerings of
   the same course}
    \label{coursecon}
\end{equation}

We interpret the linear program as follows: \begin{itemize} \item
Eqn.~\ref{qtrval}: The value of a quarter is the total number of courses taken
that quarter.  We want to maximize this value, to maximize the number of high
value classes we take each quarter.  \item Eqn.~\ref{weightlim}: In each quarter,
the total number of credits cannot exceed the specified limit.  $Weight(c_i)$
represents a short hand way of indicating the number of credits hours a class
takes.  \item Eqn.~\ref{timecon}: A student can take only one class in one particular
time slot.  Recall that we use $G_{conflict}$ to determine time conflicts.
\item Eqn.~\ref{coursecon}: Previously in
section~\ref{classDefs} we noted that it is possible for a course to have multiple
offerings during the same quarter.  A student can only take one offering
of a course during a quarter.  We build this constraint for every class-quiz section
combination for a given course.\end{itemize}

>>>>

\subsection{Representing Prerequisites} Initially, we represented course
prerequisites using a directed acyclic graph.  Each node represented a course, and edges
exist between nodes $i$ and $j$ if and only if $i$ was a prerequisite for $j$.
This model works well if each course has only one prerequisite.  However,
this model fails to represent courses that have multiple prerequisites
or courses that have alternative prerequisites.  This is the case for some courses.
We therefore concluded that it is infeasible to model prerequisites using a graph.
\begin{figure} [ht] 
    \begin{center}
        \includegraphics[scale=0.4]{prereq_logic_example} 
    \end{center}
    \caption{Evaluation of a rule} 
    \label{logic_ex} 
\end{figure}

@@

Instead of trying to solve the complete problem we chose to use the ``rolling
horizon approach" used by Xiangtong et al \cite{xiangton:informs} when
scheduling classes for airline pilots.  Xiangtong et al. also could not develop
an optimal class schedule in one pass, instead they picked optimal classes
using a heuristic over a well demarcated time period.  Once they had a solution
to the smaller time period, they increased the size of the time period, or rolled
the horizon, and used the previous optimal solution to help determine
a solution to the new larger problem.  The team's algorithm continued rolling the
time horizon on the problem until they had a solution to the entire problem.

@@

\subsection{Assumptions} Our method and in particular, our heuristic value
function assumes continuity over time in how courses are offered. Since we do
not have perfect schedule information, we don't look into the future, however we
do assume that future schedules will bare resemblance to past schedules. The
closer past course offering frequency and times offered is to the current
schedule the better our predictions will be. We also assume that a student does
not fail courses and therefore never needs to retake a course, that is once
a course is entered in the set $T$ it never leaves.

@@

\subsubsection{Procedure} We present the procedure for populating the set $L$ as
follows: \begin{enumerate} \item For each rule $r \in R$ we iterate over each
literal $p$ in the rule.  Recall that $p$ represents a course, for example
$MATH125$.  We then check if $p \in T$.  If $p \in T$, we say $p$ evaluates
to true; otherwise, $p$ evaluates to false.  Using rules of logic we evaluate
$r$.  If $r$ is true, we add the course associated with it to $L$.  Once this
is complete, $L$ will consist of every course a student is eligible to take, based solely
on satisfied prerequisites.  \item We now prune $L$ by removing all courses that
are not offered in the current quarter, and all courses which the student has
already taken. Firs we take the set difference $L-T$ and the set difference
between $L$ again and the input set of courses offered this quarter.  $L$ now
contains only classes that we can take given the current quarter and the classes
we have already taken.  \item We now construct a conflict graph, $G_{conflict}$,
where every course offering in $L$ is a node, and there is an edge between two
nodes $i$ and $j$ if and only if the time of course offering $i$ overlaps the
time of course offering $j$.  Figure~\ref{conflictg} shows an example conflict
graph for one time period.  \end{enumerate} \begin{figure} [ht] \begin{center}
\includegraphics[scale=0.35]{conflict_graph_ex} \end{center} \caption{A Conflict
Graph} \label{conflictg} \end{figure}

\subsubsection{Analysis} The selection procedure involves a lot of set queries
and iteration, but in asymptotic running time the algorithm is efficient.
Let $n$ be the total number of classes ever available to take.
In the worst case, $R$ has a complex rule for all $n$
courses, and checking each rule in $R$ is linear in the number of propositions.
However, we know {\it a priori} from analysis of our data set that the largest
rule has twelve propositions, so we can bound the rule check time by a constant,
$C$, and therefore step one runs in $O(n)$ time.

@@

Additionally, we know that not every course is offered every quarter.  If
a course is not offered until the following year, we could potentially add an
entire year to the schedule.  This is a situation we hope to avoid.  Our heuristic
function must therefore consider how often and how far in the future a given
course is offered.  For example, if we are deciding between two classes
that are identical except one is offered once a year, and the other is offered
three times a year, we want to take the course only offered once a year, as we
won't have the opportunity to take it for another entire year.
\begin{figure} [ht]
    \begin{center}
        \includegraphics[scale=0.35]{more_prereq_tree}
    \end{center}
    \caption{A Prerequisite DAG}
    \label{prereq}
\end{figure}

In
order to define our heuristic function, we must first define some auxiliary
functions on course offerings.  \begin{itemize} \item Let $TimeFromNow(c, q)$ be
difference in quarters between the current quarter, $q$, and the \emph{last} quarter
a given course, $c$, was offered.  As there are four quarters, the range
of this function is $\{0, 1, 2, 3\}$.  \item Let $Quarters(c)$ be the number of
quarters per year a given course, $c$, is offered.  As there are four quarters,
the range of this function is $\{0, 1, 2, 3\}$.  \item Let $Dependents(c)$ be
the number of course that either directly or indirectly require a given course,
$c$.  In Figure~\ref{prereq}, $Dependents(MATH125)$ = 6 while
$Dependents(MATH307)$ = 1 \end{itemize}

@@

Where $c$ is a given course offering, $q$ is the current quarter, and $\alpha$,
$\beta$, and $\gamma$ are real numbers, giving a weight to piece of the
heuristic. Determining the best weights should be done experimentally.

\subsubsection{Determining $Dependents(c)$} We briefly address how to determine
the dependent value of a particular course. Figure~\ref{prereq} indicates the
method for determining the dependent value of a course; generate the graph of
prerequisites and then count the number of children in the subtree rooted at
a particular course node. If this was the only consideration though, we could
potentially assign a very high value to a course that is not related to
a student's major because that course had a large number of dependents. The
purpose of our system is to provide schedules that get a student through a major
as quickly as possible, so we should only value courses that lead to graduation.
The set $S$ describes the classes a student will need to graduate.

To assign values to courses, we start with the courses in $S$ and work backwards
using the prerequisite rules in $R$ to determine how to traverse the
prerequisite DAG. Traversing the graph using this method, we determine the value
of the course by counting the children in the tree rooted at the current node in
the traversal. By starting with the courses in $S$ we ensure that we only given
values to courses that we need to take, instead of any courses with a large
number of dependents.

@@

\begin{equation}
   X_i + X_j \leq 1, \forall \text{ pairs } (X_i,X_j) \text{ where } X_i \text{ and } X_j \text{ are offerings of
   the same course}
    \label{coursecon}
\end{equation}

We interpret the linear program as follows: \begin{itemize} \item
Eqn.~\ref{qtrval}: The value of a quarter is the total number of courses taken
that quarter.  We want to maximize this value, to maximize the number of high
value classes we take each quarter.  \item Eqn.~\ref{weightlim}: In each quarter,
the total number of credits cannot exceed the specified limit.  $Weight(c_i)$
represents a short hand way of indicating the number of credits hours a class
takes.  \item Eqn.~\ref{timecon}: A student can take only one class in one particular
time slot, $t$  Recall that we use $G_{conflict}$ to determine time conflicts.
\item Eqn.~\ref{coursecon}: Previously in
section~\ref{classDefs} we noted that it is possible for a course to have multiple
offerings during the same quarter.  A student can only take one offering
of a course during a quarter.  We build this constraint for every class-quiz section
combination for a given course.\end{itemize}
